\clearpage
\section{Discussion}
\label{sec:discussion}

\subsection{Interpreting the Pooled Effects: Magnitude and Mechanism}

The reduced-form results in this thesis converge on a coherent, if imprecise, story. In microregions where ProUni generated a moderate inflow of college-educated workers relative to local labour demand, the formal-sector average wage rose. The most parsimonious explanation---consistent with the Card and Lemieux (\citeyear{CardLemieux2001}) framework on skilled labour supply---is that these markets were operating in a scarcity regime: private returns to tertiary credentials remained elevated because local skilled labour was relatively thin, and the programme's scholarship allocation relaxed that constraint without overloading it. In contrast, microregions that absorbed the largest volumes of ProUni scholarships by the mid-2010s display wage effects near zero. The continuous-treatment TWFE exercise cautions against reading this as pure credential inflation: once baseline household income is controlled for, the high-dose penalty weakens substantially, suggesting that the 2014--2016 recession hit richer, more credentialled labour markets hardest at exactly the moment when those microregions peaked in programme exposure. The most defensible interpretation is therefore a composition of two forces---genuine diminishing returns in the most saturated markets and a macroeconomic headwind that disproportionately loaded onto those same locations.

These estimates are measured on the average wage of \emph{all} formal workers aged 22--35 within a microregion, not on individual ProUni beneficiaries. Because beneficiaries constitute only a modest share of the local formal workforce, any aggregate effect is necessarily diluted relative to the private return earned by the marginal graduate. The thesis should therefore not be read as pinning down a precise individual return to a ProUni degree; the contribution is to show that the local reduced-form effect of educational expansion varies systematically with programme intensity---a policy-relevant result even when confidence intervals are wide, because the estimand is intentionally demanding.

\subsection{Scarcity versus Saturation: Evidence and Caution}

The contrast between Low-Dose and High-Dose microregions is consistent with a scarcity-versus-attenuation interpretation. In markets that receive a moderate inflow of graduates, wage gains are more positive; in heavily exposed markets, estimates approach zero. That pattern accords with economic theories of diminishing marginal returns to educational expansion \cite{CardLemieux2001} and potential congestion in local skilled labour markets, as also documented in the Indonesian infrastructure context \cite{Duflo2001}.

At the same time, the thesis cannot claim to have fully identified a clean causal dose-response curve. The methods chapter makes clear that standard parallel trends identifies within-stratum effects, whereas clean comparisons \emph{across} dose strata require the stronger assumption that selection into dose intensity is conditionally exogenous given observable region controls. Moreover, the continuous-treatment TWFE robustness exercise shows that the negative high-dose pattern is sensitive to controls for baseline income and macroeconomic geography. The safest formulation is therefore that the data are \emph{consistent with} saturation or attenuation in heavily exposed markets, not that they establish it conclusively.

The age-cohort decomposition adds suggestive evidence on mechanisms. In Low-Dose areas, the rise in older workers' wages is consistent with positive spillovers---improved firm-worker matching, productivity complementarities, or composition effects as higher-skilled workers displace lower-skilled workers in formal employment. In High-Dose areas, weaker young-cohort gains and negative older-cohort wage effects are consistent with relative wage compression or displacement. Yet those patterns do not uniquely identify a single structural mechanism. Because the old cohort functions simultaneously as a placebo benchmark, non-zero effects on that group signal that residual local shocks---macroeconomic downturns concentrated in high-infrastructure regions, for instance---may still be present in the data.

\subsection{Heterogeneity: Gender}
\label{sec:gender_discussion}

The gender results are the most substantively important heterogeneity finding in this thesis and merit more than a brief mention. ProUni is a majority-female programme---women account for roughly 57\% of scholarship grants across the sample period---yet the pooled formal-sector wage estimates strongly favour male workers in Low-Dose microregions (ATT $\approx$ 1.67\%) while women's estimates are near zero (ATT $\approx$ $-$0.03\%). This divergence is not a statistical artefact: it reflects a labour-market mechanism operating \emph{downstream} of educational access, precisely the kind of structural ceiling that segmented labour market theory predicts \cite{ReichGordonEdwards1973, Goldin2014}.

Three mechanisms are consistent with this pattern and are worth distinguishing. First, \textbf{occupational segregation by field of study}: \citeA{BusteloEtAl2021} document that ProUni-enrolled women disproportionately choose Health, Teaching, and Humanities programmes---sectors with compressed, often publicly regulated wage schedules. If a labour supply shock in teaching or nursing does not translate into wage premia because wages are institutionally set, the aggregate wage ATT for women will be close to zero even if women genuinely benefited from access. Second, \textbf{formal-sector selection and composition effects}: RAIS observes wages conditional on formal employment. If ProUni converts informally employed women into formally employed ones---an improvement in labour-market outcomes by any welfare measure---but these women enter formal jobs at the bottom of the formal wage distribution, the average wage in the microregion falls even as individual welfare improves. This selection-into-formality channel cannot be distinguished from a wage effect using RAIS alone \cite{EngbomEtAl2022}. Third, \textbf{labour demand constraints}: \citeA{Goldin2014} argues that the final chapter of gender wage convergence requires not merely equal credentials but equal access to nonlinear-reward jobs---managerial, professional, and client-facing roles where long, flexible hours are rewarded convexly. In the Brazilian formal sector, access to such roles is mediated by networks and firm-level hiring practices that remain demonstrably unequal \cite{Gerard2021}. ProUni removes the credential barrier but cannot dissolve the network and discrimination barriers downstream.

These three mechanisms are observationally similar in the current aggregate design. Future work that links individual ProUni records to RAIS employment spells and occupational classifications could distinguish them by testing whether women's gains are concentrated in formal-sector entry (consistent with mechanism 2) or in wage premia within pre-existing formal jobs (consistent with mechanism 1 or 3). What the present results do establish is that headline access statistics---the majority-female scholarship share---are a poor guide to the distributional incidence of wage gains. This asymmetry should be a central concern for ProUni's programme evaluation going forward.

\subsection{Heterogeneity: Ethnicity}

The ethnicity results are more encouraging than the gender results, but they also require disciplined interpretation. In moderately exposed markets, the estimates are consistent with positive wage effects for non-white workers and with a narrowing of relative disadvantage within the formal sector. That finding is meaningful given ProUni's affirmative-action design: the 50\% non-white reservation aligns programme access with the population most affected by documented labour-market discrimination \cite{Cornwell2017, AriasYamadaTejerina2002}.

Still, the thesis does not establish that ethnoracial wage inequality has disappeared, nor that full labour-market parity has been achieved. The positive ATT for non-white workers in Low-Dose regions must be weighed against two limitations. First, self-reported colour categories in RAIS are subject to reclassification over time---a phenomenon sometimes called \emph{branqueamento} (whitening), whereby workers report a lighter category as their income rises \cite[p.~86]{Telles2004}. If reclassification is correlated with programme exposure, the measured effect on the non-white wage average reflects both genuine wage gains and outflow from the non-white measurement group. Second, \citeA{Gerard2021} show that firm-level assortative matching is a primary mechanism sustaining Brazil's ethnoracial wage gap: non-white workers sort into lower-productivity firms independently of their own credentials. If ProUni does not disrupt this sorting---which seems likely given that it operates through private, non-selective HEIs---then the positive reduced-form effect is consistent with rising credentials without full convergence in firm quality.

\subsection{Geographic Concentration and the Bel\'india Problem}
\label{sec:belindia}

The geographic results reveal the central policy tension of this thesis in a precise, quantifiable form. ProUni's scholarship concentration in the South and Southeast is not incidental: it is structurally determined by the private HEI market, which tracks existing commercial demand for higher education, and by application behaviour, which reflects information access and educational preparation that are themselves geographically unequal \cite{McCowan2007}. The Bel\'india framework of Bacha \cite{bacha1976, Telles2004} predicts exactly this: market-based mechanisms, even when subsidised, tend to reinforce existing spatial hierarchies because they are calibrated to existing infrastructure rather than to social need.

The empirical operationalisation of this tension in the thesis is the North/Northeast versus South/Southeast heterogeneity result. If---as the descriptive evidence suggests---Low-Dose positive effects are disproportionately concentrated in the Centre-West, North, and Northeast, while High-Dose null effects dominate the South and Southeast, this constitutes reduced-form evidence that ProUni's returns are highest precisely where skilled labour remains scarce. The ``India'' half of Bel\'india receives a lower absolute volume of scholarships but converts each one into a larger local wage premium because the scarcity margin is steeper. Conversely, the ``Belgium'' half receives the majority of scholarships but faces diminishing marginal returns because the local skilled labour market is already relatively deep.

This inversion---highest returns where supply is lowest, lowest returns where supply is highest---motivates a concrete policy recommendation: a spatial reallocation of ProUni scholarships toward under-served microregions, conditional on the establishment of minimum HEI capacity to absorb graduates into formal employment networks. The current institutional architecture, which delegates scholarship allocation to private institutions with no spatial equity mandate, is unlikely to achieve this rebalancing without explicit policy intervention. Closer attention to regional targeting mechanisms and local absorptive capacity should therefore be a priority for future ProUni programme evaluations.

\subsection{Limitations and Statistical Precision}
\label{sec:limitations}

Several limitations qualify the interpretation of the results.

First, the analysis is restricted to the \emph{formal} labour market observed in RAIS. Informal employment, self-employment, and non-employment are not observed, so the thesis estimates effects on formal-sector outcomes rather than on total labour-market welfare. If ProUni's primary effect is to formalise workers---converting informal employees into formal ones---then the aggregate wage results mix wage and composition effects in a way that cannot be separated from wage growth using RAIS alone \cite{EngbomEtAl2022}.

Second, the treatment variable measures scholarship \emph{grants}, not observed degree \emph{completions}. Dropout and non-completion in private Brazilian HEIs---estimated at 40--57\% in INEP census data \cite{INEP2024}---introduce classical measurement error that attenuates treatment intensity and weakens the first-stage link to tertiary attainment. The near-zero first-stage estimates on the college share outcome in RAIS are consistent with this attenuation, compounded by the known under-reporting of educational credentials in RAIS \cite{EngbomEtAl2022}. As a consequence, Wald-style LATE calculations are uninformative and are not reported as primary results. The thesis remains a reduced-form analysis of local policy exposure.

Third, the estimates should be interpreted in the context of Brazil's broader higher-education policy mix. Time fixed effects absorb national shocks, but spatially heterogeneous interactions between ProUni, FIES, and the federal quota law may contaminate the estimates in High-Dose microregions where all three policies co-expanded. A direct robustness test---interacting FIES enrollment intensity from INEP HEI-level data with the post-treatment indicator---is a natural extension that would substantially strengthen the ProUni-specific interpretation of the results.
