\section{Data and Empirical Strategy}
\label{sec:methodology}

\subsection{Introduction}

This chapter details the identification strategy employed to evaluate the distributional labor market consequences of the \acrfull{prouni}. The central empirical challenge is to isolate the causal effect of a large-scale, staggered supply shock on local earnings from confounding factors such as secular growth trends or pre-existing regional divergence, especially in the context of Brazil's economic dualism.

Following the seminal framework of \citeA{Duflo2001}, I exploit two distinct dimensions of variation: the \emph{spatial intensity} of the program (the density of scholarships allocated to specific microregions) and the \emph{temporal exposure} of different cohorts (determined by the approximate timing of entry into the labor force). This interaction allows for a precise ``dosage'' design: rather than asking merely whether ProUni affected wages, this thesis investigates how the magnitude of the supply shock shapes labor market outcomes on both the extensive and intensive margins.

Departing from the parametric linear interaction models common in earlier literature, I adopt the \emph{forward-engineering} approach recommended by \citeA{BakerEtAl2025}: I first define the target causal parameter (group-time average treatment effects within dose strata), state identification assumptions explicitly, and then estimate each 2$\times$2 building block separately before aggregating transparently.\footnote{This follows the methodological framework of \citeA{CallawaySantAnna2021} for staggered binary treatments and \citeA{CallawayGoodmanBaconSantAnna2024} for continuous treatments.} Specifically, I construct a continuous, lagged measure of \acrshort{prouni} exposure (a scholarship-density ``dose''), partially aggregate this continuous dose into interpretable exposure strata, and estimate group-time average treatment effects using the \texttt{did} package of \citeA{CallawaySantAnna2021}, aggregating them into dynamic event studies. This approach relaxes the strict assumptions of constant treatment effects and linearity, allowing for the explicit detection of non-linear mechanisms such as market saturation and scarcity rents.

The chapter proceeds as follows. Section~\ref{sec:data_processing} describes the three administrative data sources and their harmonization. Section~\ref{sec:treatment_def} constructs the treatment variable. Section~\ref{sec:empirical_strategy} presents the identification assumptions and estimation equations. Section~\ref{sec:robustness} details the validation protocol.


%% ============================================================
%% DATA
%% ============================================================
\subsection{Data Sources and Processing}
\label{sec:data_processing}

The analysis integrates three primary administrative data sources, harmonized via the \texttt{basedosdados} infrastructure \cite{R-basedosdados} to ensure consistent geographic identifiers and variable definitions across the 2005--2019 period. This package sources data directly from the API of \href{https://basedosdados.org/}{Base dos Dados}, a non-profit open data platform that centralizes and harmonizes hundreds of public Brazilian datasets \cite{basedosdados.org, Dahis2022}. The data visualization in this thesis is facilitated by the \texttt{geobr} package developed by \citeA{R-geobr}.

\subsubsection{Labor Market Outcomes: \acrshort{rais}}

\paragraph{Data description.}
The primary source of the dependent variables is \acrfull{rais}, a matched employer--employee administrative census of Brazil's formal labor market created by decree in 1975. Unlike household surveys (e.g.\ \acrshort{pnad-c}), which are subject to sampling error and self-reporting bias, \acrshort{rais} provides near-universe coverage of formal earnings with geographic detail down to the municipal level \cite{rais-sumario, rais-ibge, DixCarneiro2018}. Employer reporting is mandatory, and non-compliance is punishable by fines, underpinning the high coverage of formal employment in the data \cite{EngbomEtAl2022}.

\acrshort{rais} is organized around the employment relationship (\emph{v\'inculo}): each row links a worker to an establishment in a given year and reports both worker characteristics (e.g.\ age, ethnicity/phenotype, gender) and job characteristics (e.g.\ remuneration) \cite{rais-sumario}.


\paragraph{Sample construction.}
To focus on the demographic directly exposed to the policy, I restrict the outcome sample to individuals aged \textbf{22 to 35 years}. This age band is chosen for two reasons:
\begin{enumerate}
    \item \textbf{Policy exposure:} It captures the ``first generation'' of ProUni graduates entering the workforce (typically aged 22--24 at graduation) and their early career trajectory.
    \item \textbf{Counterfactual cohort:} It excludes older cohorts whose wage dynamics are primarily driven by seniority or tenure, which might introduce noise into the estimation of returns to schooling \cite{Card1999, AngristPischke2009}.
\end{enumerate}
The panel strictly ends in 2019 to avoid the structural break caused by the COVID-19 pandemic, ensuring that the parallel trends assumption is not violated by asymmetric economic shocks in 2020.

\paragraph{Age-Cohort Decomposition.}
Following \citeA{Duflo2001}, who identified the returns to Indonesia's school construction program by comparing young cohorts (who could use the new schools) against older cohorts (who could not) within the same district, I decompose the main 22--35 sample into two age cohorts:
\begin{itemize}
    \item \textbf{Exposed cohort (ages 22--28):} Workers who could plausibly be ProUni graduates, having entered university at the standard age (18--20) and graduated within 4--6 years.
    \item \textbf{Control cohort (ages 29--35):} Workers too old to have been ProUni beneficiaries at standard university-going age during the treatment period.
\end{itemize}
Both cohorts are downloaded separately from RAIS using the same SQL structure, sample restrictions, and geographic aggregation as the main extract.


\paragraph{Main outcomes.}
The main dependent variable is the \textbf{average of log monthly remuneration} (in minimum wages) of formal workers aged 22--35. Specifically, I first compute $\ln(w_i)$ at the individual worker level within each municipality--year--demographic cell, then aggregate to the microregion--year cell by taking the population-weighted mean across municipalities within the microregion (Section~\ref{sec:unit_analysis}):
\begin{equation}
    \overline{\ln w}_{r,t} = \frac{\sum_{m \in r} N_{m,t} \cdot \overline{\ln w}_{m,t}}{\sum_{m \in r} N_{m,t}},
    \label{eq:outcome_variable}
\end{equation}
where $\overline{\ln w}_{m,t}$ is the average of individual log wages in municipality $m$ at time $t$ and $N_{m,t}$ is the number of formal workers aged 22--35 in that municipality--year cell.\footnote{This construction follows the standard Mincerian tradition in labor economics. The alternative---taking the log of the arithmetic mean wage, $\ln(\bar{w}_{r,t})$---conflates wage level effects with within-unit distributional shifts via Jensen's inequality: $\ln(\mathbb{E}[w]) \geq \mathbb{E}[\ln(w)]$, with the gap driven by within-municipality wage variance. If the treatment changes the within-unit wage distribution, the log-of-average estimand captures both level and inequality effects, complicating interpretation. The average-of-logs measures changes in the geometric mean wage, is less sensitive to outliers, and corresponds to percentage changes in individual welfare under concave utility \cite{Card1999}.} The main estimation sample comprises non-white workers (RAIS race codes 1, 4, and 8: Ind\'{i}gena, Preta, and Parda), consistent with ProUni's affirmative action mandate targeting historically disadvantaged populations. White workers (code 2, Branca) are analysed separately in the heterogeneity subsample.

In addition, I construct an extensive-margin outcome: the log total number of formal \emph{v\'inculos} of individuals aged 22--35 in each microregion--year cell, $\ln N_{r,t}$, distinguishing effects operating through wages from those through employment.

Table~\ref{tab:summary_stats} reports summary statistics for both outcomes, stratified by dose group and time period.

\subsubsection{Treatment Intensity: \acrshort{prouni}}
Treatment data is derived from the administrative records of the ProUni program, detailing the precise location (municipality of the \acrshort{hei}) and timing of every scholarship granted between 2005 and 2019.
\begin{itemize}
    \item \textbf{Spatial Aggregation:} Scholarship data are aggregated to the microregion level to align with the definition of the local labor market (see Section~\ref{sec:unit_analysis}).
    \item \textbf{Covariate Normalization:} To measure intensity rather than volume, scholarship counts are normalized by the local youth population (aged 18--24) obtained from the 2010 Demographic Census. This ensures the treatment variable reflects the ``saturation'' of the potential skilled labor force.
\end{itemize}

\subsubsection{Unit of Analysis: The Microregion}
\label{sec:unit_analysis}

While Brazil is administratively divided into 5,570 municipalities, this analysis aggregates data to the \textbf{Microregion} level ($N = 558$), an official regional division above municipalities that the \acrshort{ibge} established in 1990 according to (a) the spatial structure of industrial and agricultural production and (b) the flow of goods, capital, and information \cite{IBGE_QGR_2022}.\footnote{IBGE's meso/micro regionalization has been superseded by the Immediate and Intermediate Geographic Regions in 2017, but microregions remain a standard unit in applied work with Brazilian administrative data that reduces municipality-level commuting measurement error \cite{IBGE_QGR_2022}.}

This aggregation is necessary to address the ``spatial mismatch'' present in municipality-level data. In integrated economic zones, an individual's residential location (where the scholarship is theoretically assigned) often differs from their workplace location (where the earnings outcome is measured) due to commuting patterns. As noted in the literature on Brazilian local labor markets \cite{DixCarneiro2018}, the Microregion can be regarded as a self-contained commuting zone in which labor supply and demand equilibrate. This approach minimizes measurement error from cross-municipal commuting while maintaining 558 clusters---well above the threshold required for valid cluster-robust inference \cite{CameronMiller2015}.\footnote{The ``few clusters'' problem identified by \citeA{CameronGelbachMiller2008} arises primarily with 5--30 clusters. With 558 clusters, standard asymptotic cluster-robust standard errors are fully reliable.}

In practice, outcome variables are constructed from municipality--year RAIS aggregates and then spatially aggregated to the microregion--year level via the mapping \texttt{id\_municipio} $\rightarrow$ \texttt{id\_microrregiao}, using population weights to ensure that the microregion measure reflects the size composition of its constituent municipalities.


%% ============================================================
%% TREATMENT VARIABLE
%% ============================================================
\subsection{Defining the Treatment: The ``Dose'' of Human Capital}
\label{sec:treatment_def}

A central hypothesis of this thesis is that educational supply shocks exhibit non-linearities. A linear continuous treatment specification (e.g., ``effect per 1 additional scholarship'') assumes that the 100th scholarship has the same marginal effect as the 10,000th. This is unlikely to hold if local markets face absorptive capacity constraints. To capture potential non-linearities, I adopt a \textbf{discretized continuous treatment} approach following \citeA{CallawayGoodmanBaconSantAnna2024}.

\subsubsection{The Continuous Dose Variable ($D_{r,t}$)}
For each microregion $r$ and year $t$, I construct the cumulative scholarship density variable $D_{r,t}$:
\begin{equation}
    D_{r,t} = \frac{\sum_{k=2005}^{t-4} \text{Scholarships}_{r,k}}{\text{Youth Population}_{r, 2010}} \times 1000.
    \label{eq:dose_variable}
\end{equation}

The numerator is \textbf{lagged by four years}. Following the logic in \citeA{Duflo2001}, an educational investment in year $t$ only becomes a labor supply shock when beneficiaries plausibly complete their degree and enter the workforce. Accordingly, I interpret the lagged dose as a proxy for the local inflow of ProUni-educated workers, noting that the numerator captures \emph{scholarship grants} (not observed degree completions) and therefore represents an exposure measure rather than a literal count of graduates.

The denominator uses the 2010 Census youth population as a time-invariant normalizer, ensuring that the dose variable is not contaminated by demographic changes that may themselves be affected by the treatment.


\subsubsection{Discretization}
\label{sec:discretization}
Following the methodological recommendations of \citeA{CallawayGoodmanBaconSantAnna2024} for handling continuous treatments with high dimensionality, I partially aggregate the continuous exposure measure into a small number of empirically meaningful dose strata, yielding readable ``dose-aware'' event-study plots that retain heterogeneity across the treatment distribution.

\paragraph{Bin definition.}
I define dose bins from the \textbf{cross-sectional distribution of $D_{r,2019}$}---each microregion's lagged cumulative density in the final sample year. This reflects the full, realized treatment intensity of the program and provides maximal cross-sectional variation for distinguishing dose strata.\footnote{An alternative approach would define bins from the \emph{initial treatment year} distribution $D_{r,G_r}$, avoiding any potential look-ahead bias. In practice, the early-period distribution is extremely compressed (all microregions have near-zero density in 2009 when the lag first activates), yielding noisy bin assignments that attenuate treatment contrasts. I report initial-year bins as a sensitivity check in Section~\ref{sec:robustness}. See \citeA{AcemogluFinkelstein2008} for a related discussion on pre-treatment intensity measures.}

This choice is justified by the institutional structure of ProUni's allocation mechanism. Scholarship supply is determined by \emph{national} budget decisions and the presence of private \acrshort{hei}s willing to participate in exchange for tax exemptions. Crucially, the allocation does not respond to local labor market outcomes: MEC does not direct more scholarships to microregions with faster wage growth. The cross-sectional variation in $D_{r,2019}$ therefore largely reflects which microregions had greater \emph{pre-existing private HEI infrastructure}---a structural feature that predates the program---rather than endogenous responses to treatment effects. The quantile cutoffs from the 2019 distribution are:

\begin{itemize}
    \item \textbf{Control Group (Baseline):} Microregions whose lagged dose has not yet exceeded the 20th percentile threshold $\tau_{20}$ of the 2019 distribution ($D_{r,t} \leq \tau_{20}$). These regions received negligible exposure and serve as the counterfactual for ``no effective treatment.''
    \item \textbf{Low Dose (Scarcity):} Microregions with 2019 density between the 20th and 50th percentiles ($D_{r,2019} \in (\tau_{20}, \tau_{50}]$). These areas experienced a moderate supply shock, testing the ``Scarcity Rent'' hypothesis.
    \item \textbf{High Dose (Saturation):} Microregions with 2019 density above the 75th percentile ($D_{r,2019} > \tau_{75}$). These areas experienced a large supply shock, testing the ``Saturation'' hypothesis.
\end{itemize}

This two-stratum partition emphasizes the theoretically motivated contrast between scarcity and saturation regimes while maintaining adequate sample sizes within each bin for reliable inference. Robustness to alternative cutoff choices ($\tau_{25}/\tau_{75}$ and $\tau_{33}/\tau_{67}$) is assessed in Section~\ref{sec:robustness}.


\paragraph{Treatment-timing variable.}
The treatment-timing variable $G_d^r$ used in the staggered event-study design is defined as the \emph{first year} in which microregion $r$'s lagged density $D_{r,t}$ crosses the low-exposure threshold $\tau_{20}$:
\begin{equation}
    G_d^r = \min\bigl\{t \in [2009, 2019] : D_{r,t} > \tau_{20}\bigr\},
    \label{eq:first_treat}
\end{equation}
with $G_d^r = \infty$ for any microregion whose density never exceeds $\tau_{20}$. This ensures a consistent event-time origin $e = t - G_d^r$ for both the Low-Dose and High-Dose subsamples.


\paragraph{Comparison group.}
Since all 558 microregions eventually accumulate positive scholarship exposure over the 2005--2019 period, no microregion satisfies $D_{r,t} = 0$ permanently. The identification strategy therefore relies on \textbf{not-yet-treated} microregions as the comparison group, as explicitly sanctioned for staggered designs without a never-treated group by \citeA{CallawaySantAnna2021} (Section~5) and \citeA{CallawayGoodmanBaconSantAnna2024}. Formally, the comparison group for cohort $G_d^r = g$ at time $t$ is:
\begin{equation}
    C_t \;=\; \bigl\{r : G_d^r > t \bigr\},
    \label{eq:comparison_group}
\end{equation}
consisting of all microregions that have not yet crossed the dose threshold $\tau_{20}$ by period $t$. When estimating $ATT_d(g,t)$ for the Low-Dose subsample, units in the High-Dose bin are excluded from both the treated and the comparison sets, so that each dose-specific event study is internally coherent. This is implemented in \texttt{R} via \texttt{did::att\_gt()} with \texttt{control\_group = "notyettreated"}.


%% ============================================================
%% EMPIRICAL STRATEGY
%% ============================================================
\subsection{Empirical Strategy: Heterogeneous \acrshort{did}}
\label{sec:empirical_strategy}

Identifying the causal effect of \acrshort{prouni} faces two central econometric challenges: \emph{staggered treatment adoption} and the \emph{likelihood of heterogeneous effects across cohorts and over time}. These challenges are particularly acute for ProUni because the program rolled out first in microregions with pre-existing private \acrshort{hei} infrastructure, making early adopters structurally different from late adopters. In such settings, the standard \acrfull{twfe} estimator compares treated units not only to untreated units but also to already-treated units---so-called ``forbidden comparisons'' \cite{GoodmanBacon2021}. If treatment effects evolve over time, these comparisons receive negative weights that can severely distort estimates, potentially flipping the sign of the coefficient \cite{deChaisemartin2020, GoodmanBacon2021}. I therefore employ the non-parametric estimator proposed by \citeA{CallawaySantAnna2021} as the primary specification, benchmarking against \acrshort{twfe} as a diagnostic (Section~\ref{sec:robustness_twfe}).


\subsubsection{Identification Assumptions}
\label{sec:assumptions}

The causal interpretation of the group-time average treatment effects estimated in this framework rests on four assumptions.

\begin{description}

\item[A1 --- Parallel Trends.]
In the absence of treatment, average potential outcomes in treated and comparison microregions would have followed parallel paths. Since no microregion remains permanently untreated, I rely on the not-yet-treated parallel trends assumption. Formally, for each dose stratum $d \in \{\text{Low}, \text{High}\}$, treatment cohort $g$, and post-treatment time $t \geq g$:
\begin{equation}
    \mathbb{E}\bigl[Y_{r,t}(0) - Y_{r,g-1}(0) \mid G_d^r = g\bigr]
    =
    \mathbb{E}\bigl[Y_{r,t}(0) - Y_{r,g-1}(0) \mid G_d^r > t\bigr].
    \label{eq:pt}
\end{equation}

The plausibility of this assumption is supported by three institutional features. First, treatment timing is determined by the interaction of pre-existing private HEI infrastructure and national scholarship supply rules---not by contemporaneous local wage growth. Second, because comparison units are drawn from the same period but treated later, they share structural similarities with the treated cohort, reducing the risk of compositional bias. Third, the four-year lag in the dose variable $D_{r,t}$ (Equation~\ref{eq:dose_variable}) ensures that treatment status is predetermined with respect to current labor market conditions. The assumption is assessed empirically through pre-treatment event-study coefficients and a joint Wald test (Section~\ref{sec:robustness_pretrends}).

\item[A2 --- No Interference / SUTVA.]
The potential outcome of microregion $r$ depends only on its own treatment status and dose, not on the treatment assigned to other microregions. A violation would arise if ProUni graduates systematically migrate from treated to not-yet-treated regions, contaminating comparison-group outcomes. The use of microregions as the unit of analysis---officially defined by \acrshort{ibge} as self-contained commuting zones---substantially mitigates this risk by internalizing local labor flows. Any residual interregional migration would likely attenuate the estimated effects toward zero, meaning the results should be interpreted as conservative lower bounds.

\item[A3 --- No Anticipation.]
Microregions do not adjust their economic behavior in anticipation of future ProUni scholarship allocations. The four-year lag structure makes anticipation implausible at the microregion level: the volume of future scholarship supply depends on national budget decisions and private HEI participation choices that are not forecastable by local labor market actors.

\item[A4 --- No Confounding Contemporaneous Policies.]
No policy other than ProUni should differentially affect treated and comparison microregions in a manner correlated with treatment timing. The two major concurrent policies are \acrshort{fies} (credit expansion) and the Lei de Cotas (public university quota law). Both operate at the national level, meaning their aggregate macroeconomic impacts are absorbed by year fixed effects. However, if FIES loans expanded disproportionately in early-treated ProUni markets, a residual confound could remain. I address this interpretive concern in Section~\ref{sec:robustness_fies}.

\end{description}

\paragraph{A note on dose-response interpretation.}
A critical distinction established by \citeA{CallawayGoodmanBaconSantAnna2024} is that the standard parallel trends assumption (A1) is sufficient for identifying the \emph{level effect} $ATT(d \mid d)$---the average effect of ProUni for microregions that actually received dose $d$---but is \emph{insufficient} for causal dose-response comparisons across strata. Comparing $ATT(\text{High} \mid \text{High})$ with $ATT(\text{Low} \mid \text{Low})$ conflates the causal effect of dose with selection: microregions that receive higher doses may differ systematically in their potential treatment effect functions. Identifying clean dose-response curves requires a stronger ``strong parallel trends'' assumption that restricts treatment effect heterogeneity across dose groups.\footnote{Formally, strong parallel trends requires that the untreated potential outcome paths are parallel not just within dose groups but \emph{across} them. See Assumption~3 in \citeA{CallawayGoodmanBaconSantAnna2024}.} I report the dose-specific ATTs under standard parallel trends as the primary results and discuss dose-response patterns with the explicit caveat that they are interpretable as causal only under the stronger assumption.


\subsubsection{Estimation}
\label{sec:estimation_equations}

For each intensity bin $d \in \{\text{Low}, \text{High}\}$, I estimate group--time average treatment effects, $ATT_d(g,t)$, which compute the average effect for microregions first treated in year $g$ at time $t$, relative to the not-yet-treated comparison group $C_t$.

\paragraph{Outcome 1: Log Wages (Intensive Margin).}
The primary specification, estimated via \texttt{did::att\_gt()}, targets:
\begin{equation}
    ATT_d^w(g, t) = \mathbb{E}\bigl[\overline{\ln w}_{r,t}(1) - \overline{\ln w}_{r,t}(0) \mid G_d^r = g\bigr],
    \label{eq:main_wages}
\end{equation}
where $\overline{\ln w}_{r,t}$ is the population-weighted average of individual log monthly remuneration (in minimum wages) for formal workers aged 22--35 in microregion $r$ and year $t$ (Equation~\ref{eq:outcome_variable}). Standard errors are clustered at the microregion level using the multiplier bootstrap (see below).

\paragraph{Outcome 2: Formal Employment Count (Extensive Margin).}
An analogous specification targets the extensive-margin effect:
\begin{equation}
    ATT_d^N(g, t) = \mathbb{E}\bigl[\ln N_{r,t}(1) - \ln N_{r,t}(0) \mid G_d^r = g\bigr],
    \label{eq:main_employment}
\end{equation}
where $\ln N_{r,t}$ is the log number of formal \emph{v\'inculos} among individuals aged 22--35 in microregion $r$ at time $t$.

\paragraph{Dynamic Aggregation (Event Studies).}
To assess the dynamic evolution of the effect and visually test Assumption~A1, the $ATT_d(g,t)$ parameters are aggregated into an event-study framework relative to the treatment year $g$:
\begin{equation}
    ATT_{ES}^d(e) = \sum_{g} w_g \, ATT_d(g, g + e),
    \label{eq:event_study}
\end{equation}
where $e = t - g$ is event time and $w_g$ are cohort-size weights. This is implemented via \texttt{did::aggte(type = "dynamic")}.

\paragraph{Outcome 3: Cohort Wage Gap (Triple-Difference).}
The primary outcome of the age-cohort analysis is the within-microregion wage gap between the young and old cohorts:
\begin{equation}
    \Delta w_{r,t} = \bar{w}_{r,t}^{\text{young}} - \bar{w}_{r,t}^{\text{old}},
    \label{eq:wage_gap}
\end{equation}
where $\bar{w}_{r,t}^{c}$ is the average log wage for cohort $c$ in microregion $r$ at time $t$. Any local shock that affects all age groups equally---such as a factory closure, commodity price change, or macroeconomic fluctuation---cancels in the differencing, leaving only the age-specific ProUni channel. This yields an estimating equation analogous to Equation~\ref{eq:main_wages}:
\begin{equation}
    \Delta w_{r,t}
    \;=\;
    ATT_d^{\Delta}(g, t)
    \;\cdot\;
    \mathbf{1}\{G_d^r = g,\; T = t\}
    \;+\;
    \alpha_r + \lambda_t + \varepsilon_{r,t},
    \label{eq:main_gap}
\end{equation}
where a positive $ATT_d^{\Delta}$ indicates that ProUni raised the relative wage of the exposed (young) cohort. Additionally, estimating Equations~\ref{eq:main_wages} separately for each age group provides a decomposition: if the young-only ATT is positive and the old-only ATT is zero, the effect is confirmed as age-specific. If both are positive, positive general equilibrium spillovers are present.

\paragraph{Outcome 4: First Stage --- College Attainment Gap.}
To recover the implied return to a ProUni degree, I estimate the effect of scholarship density on the college attainment gap:
\begin{equation}
    \Delta s_{r,t}
    \;=\;
    ATT_d^{s}(g, t)
    \;\cdot\;
    \mathbf{1}\{G_d^r = g,\; T = t\}
    \;+\;
    \alpha_r + \lambda_t + \varepsilon_{r,t},
    \label{eq:first_stage}
\end{equation}
where $\Delta s_{r,t} = s_{r,t}^{\text{young}} - s_{r,t}^{\text{old}}$ is the difference in college-educated worker shares between cohorts. The college-educated share is computed from the RAIS education variable (\texttt{grau\_instrucao\_apos\_2005} $\geq 9$, capturing workers with completed tertiary education or above). The implied Wald/LATE estimate of the return to a ProUni degree is then:
\begin{equation}
    \hat{\rho}_{\text{LATE}}
    \;=\;
    \frac{\widehat{ATT}_d^{\Delta}}{\widehat{ATT}_d^{s}},
    \label{eq:wald}
\end{equation}
with standard errors computed via the delta method. This Wald estimator identifies the local average treatment effect (LATE) for compliers---students who obtained a degree because of ProUni but would not have done so otherwise---under a monotonicity assumption: higher scholarship density weakly increases the probability of college completion for all types of students in the exposed cohort. The Wald estimate should be interpreted cautiously given the employer-reported nature of the RAIS education variable, which introduces measurement error that attenuates the first stage.


\paragraph{Heterogeneity Subsamples.}
To investigate distributional impacts, Equations~\ref{eq:main_wages} and \ref{eq:main_employment} are re-estimated on mutually exclusive demographic subsamples defined by Gender (Female vs.\ Male), Ethnicity (Non-white vs.\ White), and Region (North/Northeast vs.\ South/Southeast/Centre-West). For each subgroup, the outcome variable is computed strictly from the corresponding subset of RAIS records, maintaining the same cohort definitions and comparison groups.


\subsubsection{Inference}
\label{sec:inference}

All confidence intervals are constructed using the \textbf{multiplier bootstrap} implemented in the \texttt{did} package with \texttt{bstrap = TRUE} and \texttt{cband = TRUE} (the package defaults). This procedure produces \emph{simultaneous} confidence bands that cover the entire path of event-study coefficients with probability $(1 - \alpha)$, accounting for the multiple testing problem inherent in evaluating many group-time ATTs simultaneously.\footnote{The multiplier bootstrap uses weights drawn from the influence function representation of the estimator and produces critical values larger than 1.96 (typically $\sim$2.7), reflecting the simultaneous coverage requirement. Pointwise confidence intervals (setting \texttt{cband = FALSE}) would have 95\% marginal coverage but much lower joint coverage and are therefore not reported as the primary inference tool \cite{CallawaySantAnna2021}.} Standard errors are clustered at the microregion level, which is the level of treatment assignment.


\subsubsection{Mechanisms: Testing for General Equilibrium Effects}
\label{sec:mechanisms}

A key theoretical concern in large-scale education programs is the potential for general equilibrium effects to dampen private returns. As noted by \citeA{Duflo2001}, a massive supply shock could depress the wage premium through oversupply. This thesis tests this hypothesis by comparing the dose-specific event studies:
\begin{itemize}
    \item \textbf{Scarcity hypothesis:} If $ATT_{ES}^{\text{Low}}(e) > ATT_{ES}^{\text{High}}(e)$ for post-treatment event times, the wage premium is preserved only when the supply shock is relatively scarce.
    \item \textbf{Saturation hypothesis:} If $ATT_{ES}^{\text{High}}(e)$ is significantly lower or statistically indistinguishable from zero, it indicates that massive aggregate supply shocks erode private returns through credential inflation and labor market congestion.
\end{itemize}

As discussed above, these cross-dose comparisons are interpretable as causal dose-response effects only under the strong parallel trends assumption. Under standard parallel trends, they reflect a combination of causal dose effects and selection into dose levels.


%% ============================================================
%% ROBUSTNESS
%% ============================================================
\subsection{Robustness Checks and Validity Tests}
\label{sec:robustness}

The following validation protocol is executed in Section~\ref{sec:results} to systematically test the identification assumptions and probe the sensitivity of the main findings to specification choices.



\subsubsection{Testing Assumption A1: Pre-Trend Analysis}
\label{sec:robustness_pretrends}

The primary diagnostic for the parallel trends assumption is the inspection of pre-treatment event-study coefficients $ATT_{ES}^d(e)$ for $e < 0$. Under A1, these should be statistically indistinguishable from zero. I report both point estimates with simultaneous confidence bands and a joint Wald test for pre-trends (the null hypothesis being that all pre-treatment dynamic effects are jointly zero).

\subsubsection{Age-Cohort Placebo}
As a built-in falsification test, the main specification is estimated separately for the young (22--28) and old (29--35) cohorts. Under the identifying assumption that ProUni's human capital channel operates exclusively through the exposed cohort, the old-cohort ATT should be zero. A non-zero old-cohort effect would indicate either (a) a violation of the identifying assumption (local economic shocks correlated with treatment timing) or (b) positive general equilibrium spillovers from the inflow of college-educated workers. The latter interpretation is distinguished from the former by examining whether the old-cohort effect is present only in treated microregions and only after treatment onset.

This within-market placebo is more powerful than an external age-cohort test (e.g., workers aged 36--50) because the old cohort inhabits the \emph{same microregion-year cell} as the young cohort, perfectly controlling for local conditions. Any remaining differential between young and old must therefore arise from age-specific mechanisms rather than shared local trends.

\subsubsection{Implied Returns via Wald LATE}
As described in Section~\ref{sec:estimation_equations}, the ratio of the reduced-form wage gap ATT to the first-stage college share ATT yields an implied LATE estimate. This is compared to Mincerian benchmarks from the Brazilian literature \cite{BarbosaFilhoPessoa2008, MenezesFilho2001, Card1999} to assess whether the magnitudes are economically plausible. Given the likely attenuation bias in the RAIS first stage, the Wald estimate should be interpreted as an upper bound on the true LATE.

\subsubsection{Testing for Omitted Variable Bias: Continuous TWFE and Doubly Robust Covariates}
\label{sec:robustness_dr}

Given the explicit geographic concentration of ProUni in Brazil's wealthier South and Southeast, there is a risk that simple group-time estimators confound the causal effect of ProUni with regional macroeconomic shocks (e.g., the 2014--2016 Brazilian recession). To test whether treatment is confounded by pre-existing regional wealth, I employ two strategies. First, I estimate a sequence of continuous treatment TWFE models (Linear, Logarithmic, and Quadratic doses) interacting the continuous $D_{r,t}$ with a post-treatment dummy to detect non-linear saturation patterns. 

Second, to address omitted variable bias, I estimate a covariate-adjusted TWFE model and a Doubly-Robust Callaway \& Sant'Anna specification. These models interact the post-treatment dummy with baseline (2010 Census) microregion characteristics, notably the local youth population, the share of non-white workers, and the average household income per capita. If the inclusion of these covariates---particularly baseline wealth---absorbs the negative effects observed in High-Dose microregions, it proves that the wage penalties are partially driven by macroeconomic contractions disproportionately hitting wealthy geographic hubs rather than pure credential saturation.


\subsubsection{Placebo Cohort Test}
\label{sec:robustness_placebo}

As a falsification exercise, I re-estimate the main wage models on the sample of formal workers aged \textbf{36--50 years}---a cohort too old to have been directly exposed to ProUni at standard university-going age. If the main results capture a genuine human capital supply shock from ProUni graduates entering the labor market, this older cohort should show no significant post-treatment effects. Conversely, non-zero effects in the placebo cohort would indicate that the baseline results reflect broad local economic shocks correlated with treatment timing rather than the specific mechanism of skill supply expansion.


\subsubsection{Discretization Sensitivity}
\label{sec:robustness_discretization}

To ensure results are not driven by the chosen dose thresholds, I re-run the main wage models under two alternative partition schemes: (i) $\tau_{25}/\tau_{75}$, which narrows the Low bin and sharpens the treatment--comparison contrast; and (ii) $\tau_{33}/\tau_{67}$, which provides a symmetric tercile split with more balanced bin sizes.


\subsubsection{Alternative Lag Specification}
\label{sec:robustness_lag}

The baseline specification assumes a four-year lag between scholarship allocation and labor market entry. As a sensitivity check, I re-estimate all models with a \textbf{three-year lag}, which assumes faster degree completion. The lag sensitivity test serves a dual purpose: it probes the plausibility of the timing assumption and provides indirect evidence on whether results are sensitive to changes in the effective treatment window.


\subsubsection{TWFE Benchmark}
\label{sec:robustness_twfe}

To illustrate the empirical relevance of heterogeneous treatment effect bias, I estimate the naive \acrshort{twfe} specification:
\begin{equation}
    \overline{\ln w}_{r,t}
    \;=\;
    \beta_{\text{TWFE}} \cdot \mathbf{1}\{D_{r,t} > \tau_{20}\}
    \;+\;
    \alpha_r + \lambda_t + \varepsilon_{r,t},
    \label{eq:twfe_benchmark}
\end{equation}
with standard errors clustered at the microregion level, implemented via \texttt{fixest::feols()}. As shown by \citeA{GoodmanBacon2021}, $\hat{\beta}_{\text{TWFE}}$ is a weighted average of all 2$\times$2 comparisons---including ``forbidden'' comparisons where already-treated units serve as controls. If treatment effects are heterogeneous across cohorts or evolve over time, some of these weights can be negative, attenuating or sign-reversing the TWFE estimate relative to the properly aggregated Callaway--Sant'Anna ATTs. A substantial divergence between the two estimates validates the methodological choice of the heterogeneity-robust estimator.


\subsubsection{Concurrent Policy Controls}
\label{sec:robustness_fies}

\acrshort{fies} and the Federal Quota Law (Lei de Cotas) constitute important concurrent shocks in the higher education market. Since both policies are national in scope and are not designed to target microregions based on their ProUni scholarship density, their common impact is absorbed by year fixed effects and does not mechanically confound the within-microregion event-time variation. Nonetheless, their presence motivates a cautious interpretation of the estimated effects, which are best seen as capturing the net contribution of ProUni within a broader policy mix that simultaneously expanded higher education access through complementary channels.