\clearpage
\section{Literature Review}
\subsection{Human Capital and Segmented Labor Market Theories}

Returns to education and human capital have been subject to multiple empirical and theoretical studies among economists. One of the first papers to recognize the potential interdependence between education and income distributions was \shortciteA{Mincer1958}, who later established a relation between the log of individual earnings and schooling  in which education appears as a parameter entering the equation linearly \cite{Mincer1974}.

According to the \emph{human capital} and \emph{segmented labor market} theories, returns can, however, differ when the same characteristics interact with discrimination, networks, and local demand \cite{Becker1964, Mincer1974}. In his seminal work, \citeA{Becker1964} shows that the rate of return to college education vary by gender, ethnicity, and region, i.e., rural/urban divide, in a systematic way even after adjusting for ability and opportunity costs. The earnings differential in white male cohorts between college and high-school graduates is shown to be larger than cohorts from other demographic backgrounds. Nevertheless, he concludes that investment in education remains profitable for these underrepresented groups. The findings established one of the first empirical baseline estimates for human capital theory and imply that unobserved factors, rather than ability, may drive much of the observed heterogeneity in educational returns\footnote{c.f. Chapter 4 of \citeA{Becker1964}}.

In another major empirical study conducted on the effects of education on earnings, \citeA{Card1999} revisited these results and confirmed their positive direction, as predicted by \citeA{Mincer1974}'s human capital earnings function, with the use of causal inference methods, although the magnitude is slightly higher than previous OLS estimates. Precisely, he instrumented proximity to colleges on schooling years to estimate the impact of wages, hence addressing endogeneity problems by exploiting variation in access across U.S. regions. \citeA{Card1999} found that each extra year of education led to about a 7-10\% increase in earnings. He explains that comparing siblings can give results similar to or even lower than OLS because while it lowers bias from ability, it increases errors in measurement. He also warns that using family background as an instrument can inflate returns compared to OLS. Policy-based instruments, like school changes or aid rules, usually give estimates about 20\% higher than OLS, which may be because those affected by the policies get more out of the extra schooling. Finally, \citeA{Card1999} highlights sources of heterogeneity in returns, i.e. parental education, school quality and location, while noting that "sheepskin" effects may contribute in tandem with productivity gains. This effect describes the empirical observation first made by \citeA{Hungerford1987} that higher education years imply higher educational returns which signals gains caused by the status of a diploma itself. In this regard, \citeA{BelmanHeywood1991} sought to deconstruct the sheepskin effects by looking into their impact along ethnic and gender dimensions.
\paragraph{}
While \emph{human capital} theory interprets these wage differentials as the result of individual investment outcomes in competitive markets, \emph{segmented labor market} theory complements this view by emphasizing institutional factors, hierarchies, and inequalities in access to high-return markets \cite{ReichGordonEdwards1973,DoeringerPiore1971,Leontaridi1998}. The theory conceptualized in \citeA{ReichGordonEdwards1973}'s paper argues that the labor market is divided into a primary and secondary sector in a systemic way that has been established in America due to endogenous forces. Moreover, they see segmentation as functional and not accidental, thus reinforcing the stability of capitalist institutions. 

\citeA{Leontaridi1998} offers a synthesis of segmented labor market theory. She recognizes its value in understanding the institutional and social origins of wage inequality but points out its theoretical and empirical weaknesses. \citeA{Leontaridi1998} suggests that the theory lacks a clear analytical structure, which leads to inconsistent results due to different definitions of segments and research methods. Although segmentation is a useful descriptive idea, its explanatory power needs to be combined with other ideas, such as human capital and search models. This suggests a need for studies that measure how much structural barriers, instead of productivity differences, cause differences in educational returns.
\paragraph{}
The conclusions of the two theories make it important to study the institutional mechanisms that drive the apparent heterogeneity in educational returns and come with direct policy implications. If segmentation drives return differentials along demographic lines, then education policies solely aimed at expanding access may have a limited effect on reducing inequality or lifting people out of poverty. 

Therefore, I expect \textbf{ greater effects} where policies open doors to selective programs and dense networks, thus lowering ethnic and gender gaps, and \textbf{ smaller effects} where regional labor demand limits the conversion of education into high-quality jobs. Since most of the aforementioned studies were isolated to the U.S. labor market, Section \ref{subsec:international_evidence} provides an overview of comparable empirical evidence in international contexts. 




\subsection{Returns to Education in Brazil}
The conclusions of the two theories make it important to study the institutional mechanisms that drive the apparent heterogeneity in educational returns and come with direct policy implications. If segmentation drives return differentials along demographic lines, then education policies solely aimed at expanding access may have a limited effect on reducing inequality or lifting people out of poverty.
The analysis of educational returns in Brazil must therefore be contextualized within the country's profound structural duality in all dimensions of development, famously termed \textbf{\emph{Belíndia}} by the economist Edmar Bacha in the 1970s \cite[p.~19]{Telles2004}. This blended word highlights the coexistence of a ``small, growing industrial state such as Belgium`` alongside a ``great mass of poor stagnating at almost the level of rural India`` \cite{bacha1976}. The entrenched economic segmentation justifies why educational policies must simultaneously address individual human capital deficits and the deep structural barriers that restrict social mobility between these two ``countries``. Rather than treating economic agents as a homogeneous entity, this framework naturally leads to an analysis of the specific racial, gender, and regional market failures observed in Brazil.
In the context mentioned above, several contributions to Brazilian social science employed methodologies in the tradition of \citeA{Mincer1974} to estimate educational returns with a pronounced focus on heterogeneity.
\paragraph{}
First, evidence shows ethnic wage gaps beyond observables and a skin color gradient, which has been highlighted in \citeA{AriasYamadaTejerina2002} and a seminal book on racial inequality written by \citeA{Telles2004}. Specifically, \citeA[p.~171]{Telles2004} points out a disproportionately high disparity in the income distribution only comparable to South Africa, considering other major countries with similar degrees of economic development. In his meta-study of racial discrimination, \citeA[pp.~156-163]{Telles2004} presents several experiments that highlight the existence of a glass-ceiling mechanism separating whites from nonwhites, which hinders the latter group almost completely from reaching mobility into the middle class. In particular, before the introduction of education policies, the wage gap beyond observable characteristics in Brazil was shown to be of fairly high magnitude. \citeA{AriasYamadaTejerina2002} demonstrate an unexplained difference of 18\% between non-white university graduates and their white peers based on data roughly covering a time period from 1940 to 1990 using pupil-teacher ratios.
\paragraph{}
Second, there is evidence on the existence of gender gaps that persist even though women have equal or a higher level of education, partly due to self-selection into majors and occupations as well as systemic glass-ceiling dynamics \cite{GarciaNopoSalardi2009, BusteloEtAl2021}...
\paragraph{}
Third, there are sizable regional differences connected to weaker labor demand in the North and Northeast regions as examined thoroughly by \citeA{deoliveira2021}. Shockingly, the average income from labor in the Northeastern Fortaleza area is more than 60\% lower than the national average of all Brazilian metropolitan regions \cite{deoliveira2021}. In addition, they demonstrate a positive gradient in these differences between men and women. This raises a significant challenge to develop equitable policy tools in an attempt to mitigate these differences through general education policies, such as expanding access to university for marginalized populations.
\paragraph{}
Policies can reshape these margins \emph{unevenly}. Creating equal access to high-quality education is shown to enhance educational outcomes as previously discussed but the key factor for more equitable labor market rents is access to high-quality jobs. \citeA{RiehlMachadoReyes2023} scrutinized the direct and spillover effects of affirmative action quotas for marginally admitted students of African descent from low-income households in an elite public university, UERJ, employing a two-stage regression discontinuity design. They find up to a 14\% increase in the earnings of beneficiaries six to nine years after UERJ application which are largely attributable through network access to high-paying firms. However these gains are shown to be decreasing in career progression which alludes to structural barriers in labor markets. The spillover effect of this large-scale affirmative action is revealed to impact the top enrollees of UERJ including the affiliated business school, URM, negatively. In majors with high take-up rates, early career earnings are estimated to decrease by 14\% for this subgroup. Their results also imply greater gains where networks matter, e.g. in the productive and densely populated metropolitan area of São Paulo and smaller gains where local job ladders are thin, e.g. in the Northeast of Brazil.



\subsection{International Evidence on Education Policies} \label{subsec:international_evidence}

The quantification and evaluation of policy measures related to schooling is empirically well studied in academia. One of the most extensive investigations of global returns was conducted by \citeA{PsacharopoulosPatrinos2004} in a meta-study of 1,120 estimates in 139 countries.
\\
There are several unique studies investigating different types of investment in schooling. One of them concerns the educational institution itself. 
In a widely cited quasi-experiment, \shortciteA{Duflo2001}'s studies the impact of Indonesia's school construction expansion in the 1970s and the associated effects on schooling years and labor income. In this particular article, she exploited the spatial variation in the intensity of school construction to estimate the outcomes of the labor market within the affected cohorts. \shortciteA{Duflo2001} estimated an annualized return to education of up to 10\%, which is quite substantial for a growing economy and seems to be in line with previous estimates suggested by human capital theory.
\\
Evidence from other settings reinforces heterogeneous policy returns: India’s reserved seats altered college quality and occupational sorting for disadvantaged groups \cite{BagdeEppleTaylor2016}, while U.S. affirmative action affects field choice, completion, and earnings in ways that vary by institution selectivity and demographics \cite{Bleemer2022, ArcidiaconoLovenheim2020}. 